---
title: "Ractor local GC ruby-bench A/B performance analysis"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    theme: darkly
    toc: true
    toc_float:
      collapsed: false
    code_download: true
params:
  ab_file: "data/output_054.json"
  aa_file: "data/output_053.json"
  min_effect: 0.01
  drift_threshold: 0.03
  top_n: 6
---

```{css, echo=FALSE}
:root {
  --bg: #262b37;      --card: #2e3444;   --card2: #363d50;  --border: #434c61;
  --text: #dfe4ef;    --muted: #a0aabf;  --faint: #8690a6;
  --red: #eba8a0;     --red-soft: rgba(235,168,160,.13);
  --green: #9ed4ae;   --green-soft: rgba(158,212,174,.13);
  --amber: #e7c98d;   --amber-soft: rgba(231,201,141,.13);
  --blue: #a3bfea;    --blue-soft: rgba(163,191,234,.14);
}
body { background: var(--bg); color: var(--text);
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Inter, Roboto, sans-serif;
  font-size: 16.5px; line-height: 1.6; }
.main-container { max-width: 1150px; }
h1.title { font-size: 2rem; font-weight: 800; letter-spacing: -0.02em; color: #eef1f8; }
h4.date { color: var(--faint); font-size: 0.9rem; }
h1, h2, h3 { color: #eef1f8; font-weight: 700; letter-spacing: -0.01em; }
h1 { font-size: 1.5rem; border-bottom: 1px solid var(--border); padding-bottom: 7px; margin-top: 2.3rem; }
h2 { font-size: 1.2rem; margin-top: 1.7rem; }
h3 { font-size: 1.05rem; }
a { color: var(--blue); }
p, li { color: var(--text); }
strong { color: #f5f7fc; }
code { background: #3a4155; color: #c7d0e2; border-radius: 4px; padding: 1px 5px; font-size: 0.85em; }
pre, pre code { background: var(--card2) !important; color: #c2cbde; border: 1px solid var(--border);
  border-radius: 8px; font-size: 0.82em; }

/* ---- banners ---- */
.banner { border: 1px solid var(--border); border-radius: 10px; padding: 13px 17px;
  margin: 10px 0; background: var(--card); line-height: 1.6; }
.banner b, .banner strong { color: #f5f7fc; }
.banner.good  { border-color: rgba(158,212,174,.4); background: var(--green-soft); }
.banner.bad   { border-color: rgba(235,168,160,.4); background: var(--red-soft); }
.banner.info  { border-color: rgba(163,191,234,.4); background: var(--blue-soft); }
.banner.warn  { border-color: rgba(231,201,141,.4); background: var(--amber-soft); }

/* ---- chips ---- */
.chip { display: inline-block; border-radius: 999px; padding: 2px 11px; font-size: 0.8rem;
  font-weight: 600; margin: 0 3px; white-space: nowrap; }
.chip-blue { background: var(--blue-soft);  color: var(--blue);  border: 1px solid rgba(163,191,234,.4); }
.chip-red  { background: var(--red-soft);   color: var(--red);   border: 1px solid rgba(235,168,160,.4); }
.chip-grey { background: rgba(160,170,191,.12); color: var(--muted); border: 1px solid var(--border); }

/* ---- at-a-glance tables ---- */
.gtable { font-size: 0.88rem; margin: 10px 0 22px 0; }
.gtable td, .gtable th { padding: 5px 12px; }
.gtable .gname { font-family: ui-monospace, SFMono-Regular, Menlo, Consolas, monospace;
  font-size: 0.84rem; white-space: nowrap; }
.gtable .gchg { font-weight: 800; text-align: right; white-space: nowrap; font-size: 0.95rem; }
.gtable th.r { text-align: right; }
.gtable .gbar { width: 20%; min-width: 100px; }
.gfill { height: 10px; border-radius: 3px; min-width: 2px; }
.gfill.reg { background: linear-gradient(90deg, rgba(235,168,160,.95), rgba(235,168,160,.5)); }
.gfill.imp { background: linear-gradient(90deg, rgba(158,212,174,.95), rgba(158,212,174,.5)); }
.gfill.sug { background: linear-gradient(90deg, rgba(231,201,141,.95), rgba(231,201,141,.5)); }
.gtable .gtime, .gtable .gci { color: var(--muted); white-space: nowrap; }
.gtable .gaa { color: var(--faint); white-space: nowrap; }
.gdrift { color: var(--amber); font-weight: 700; }

/* ---- tables ---- */
table { width: 100%; border-collapse: collapse; margin: 14px 0; font-size: 0.9rem;
  background: var(--card); border: 1px solid var(--border); border-radius: 10px; overflow: hidden; }
th { color: var(--muted); font-size: 0.74rem; text-transform: uppercase; letter-spacing: 0.06em;
  padding: 10px 13px; border-bottom: 1px solid var(--border); background: var(--card2);
  text-align: left; }
td { padding: 8px 13px; border-bottom: 1px solid rgba(67,76,97,.55); color: var(--text); }
tr:last-child td { border-bottom: none; }
tr:hover td { background: var(--blue-soft); }
td.pos, span.pos { color: var(--red);   font-weight: 700; }
td.neg, span.neg { color: var(--green); font-weight: 700; }
caption { color: var(--faint); font-size: 0.83rem; caption-side: top; padding: 4px 0; }

/* ---- details / summary ---- */
details { border: 1px solid var(--border); border-radius: 10px; padding: 11px 17px;
  margin: 12px 0; background: var(--card); }
summary { cursor: pointer; color: var(--blue); font-weight: 600; }
details[open] summary { margin-bottom: 8px; }

/* ---- floating TOC ---- */
.tocify { background: var(--card); border: 1px solid var(--border); border-radius: 10px; }
.tocify .tocify-item, .tocify .list-group-item {
  color: var(--muted); font-size: 0.88rem; background: transparent; border: none;
  padding: 6px 14px; cursor: pointer; }
.tocify .tocify-item:hover { background: var(--blue-soft); color: var(--text); }
.tocify .tocify-item.active, .tocify .list-group-item.active {
  background: var(--blue-soft) !important; color: var(--blue) !important;
  font-weight: 600; }

img { border-radius: 10px; border: 1px solid var(--border); max-width: 100%;
  box-shadow: 0 1px 4px rgba(0, 0, 0, 0.25); }
.figure, .caption { color: var(--faint); }
#TOC { z-index: 10; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 10, dev = "png", dpi = 130,
                      dev.args = list(bg = "#2e3444"))
library(jsonlite); library(dplyr); library(tidyr); library(purrr); library(tibble)
library(stringr); library(ggplot2); library(scales); library(knitr)

C_BG <- "#2e3444"; C_GRID <- "#3c4457"; C_TEXT <- "#dfe4ef"; C_MUTED <- "#9ba5ba"
PAL <- c(regression = "#eba8a0", improvement = "#9ed4ae", inconclusive = "#707c95")
C_BAND <- "#3d465c"; C_ZERO <- "#5b667f"; C_MARK <- "#8b96ac"
C_SIDE_BASE <- "#97a2b8"; C_SIDE_CAND <- "#a3bfea"

theme_report <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(plot.background = element_rect(fill = C_BG, color = NA),
          panel.background = element_rect(fill = C_BG, color = NA),
          panel.grid.major = element_line(color = C_GRID, linewidth = 0.45),
          panel.grid.minor = element_blank(),
          text = element_text(color = C_TEXT),
          axis.text = element_text(color = C_MUTED),
          axis.title = element_text(color = C_MUTED, size = rel(0.85)),
          plot.title = element_text(color = "#eef1f8", face = "bold", size = rel(1.1)),
          plot.subtitle = element_text(color = C_MUTED, size = rel(0.8),
                                       margin = margin(b = 10)),
          legend.text = element_text(color = C_TEXT),
          legend.position = "top", legend.justification = "left",
          strip.text = element_text(color = C_TEXT, face = "bold"),
          plot.margin = margin(14, 16, 10, 10))
}
theme_set(theme_report())

fmt_ratio <- function(x) sprintf("%.3f", x)
fmt_x     <- function(x) sprintf("%.2f\u00d7", x)
fmt_chg   <- function(r) sprintf("%+.1f%%", (r - 1) * 100)
fmt_ms <- function(s) {
  ms <- s * 1000
  dplyr::case_when(ms >= 100 ~ sprintf("%.0f ms", ms),
                   ms >= 10  ~ sprintf("%.1f ms", ms),
                   TRUE      ~ sprintf("%.2f ms", ms))
}
pct_lab <- function(b) sprintf("%+.0f%%", (b - 1) * 100)
geomean <- function(x) exp(mean(log(x)))

# "ruby 4.1.0dev (2026-08-05T02:35:23Z ractor-local-gc 1ec37f2d87) +ZJIT +PRISM [x86_64-linux]"
parse_build <- function(desc) {
  m <- str_match(desc, "^(\\S+ \\S+) \\((\\S+) (\\S+) ([0-9a-f]+)\\) (.*)$")
  if (is.na(m[1, 1]))
    return(tibble(version = desc, committed = NA, branch = NA, commit = NA, platform = NA))
  tibble(version = m[1, 2], committed = m[1, 3], branch = m[1, 4],
         commit = m[1, 5], platform = m[1, 6])
}
short_build <- function(desc) {
  p <- parse_build(desc)
  if (is.na(p$commit)) return(desc)
  paste0(p$branch, " @ ", p$commit)
}
```

```{r functions}
load_run <- function(path) {
  j <- jsonlite::fromJSON(path, simplifyVector = TRUE, simplifyDataFrame = FALSE)
  sides <- names(j$raw_data)
  stopifnot(length(sides) == 2)
  iters <- map_dfr(sides, function(s) {
    imap_dfr(j$raw_data[[s]], function(b, name) {
      tibble(side = s, benchmark = name,
             iter = seq_along(b$bench), time = as.numeric(b$bench))
    })
  })
  meta <- map_dfr(sides, function(s) {
    imap_dfr(j$raw_data[[s]], function(b, name) {
      tibble(side = s, benchmark = name,
             rss = as.numeric(b$rss %||% NA_real_),
             maxrss = as.numeric(b$maxrss %||% NA_real_),
             n_warmup = length(b$warmup), n_bench = length(b$bench))
    })
  })
  cmds <- unlist(map(j$raw_data[[sides[1]]], "command_line"))
  list(path = path, sides = sides, desc = j$metadata, iters = iters, meta = meta,
       example_cmd = if (length(cmds)) cmds[[1]] else NA_character_)
}

# ratio > 1: candidate slower. CI: delta method, conservative df = min(n) - 1.
summarise_pair <- function(run) {
  base <- run$sides[1]; cand <- run$sides[2]
  per_side <- run$iters |>
    arrange(benchmark, side, iter) |>
    group_by(benchmark, side) |>
    summarise(n = n(), mu = mean(time), s = sd(time),
              drift = mean(time[iter > max(iter) / 2]) / mean(time[iter <= max(iter) / 2]),
              .groups = "drop")
  per_side |>
    pivot_wider(names_from = side, values_from = c(n, mu, s, drift)) |>
    rename(n_b = !!paste0("n_", base),  n_c = !!paste0("n_", cand),
           mu_b = !!paste0("mu_", base), mu_c = !!paste0("mu_", cand),
           s_b = !!paste0("s_", base),   s_c = !!paste0("s_", cand),
           drift_b = !!paste0("drift_", base), drift_c = !!paste0("drift_", cand)) |>
    filter(!is.na(mu_b), !is.na(mu_c)) |>
    mutate(ratio = mu_c / mu_b,
           se = ratio * sqrt((s_b / mu_b)^2 / n_b + (s_c / mu_c)^2 / n_c),
           crit = qt(0.975, df = pmax(pmin(n_b, n_c) - 1, 2)),
           lo = ratio - crit * se, hi = ratio + crit * se)
}

rss_pair <- function(run) {
  base <- run$sides[1]; cand <- run$sides[2]
  run$meta |>
    select(benchmark, side, rss, maxrss) |>
    pivot_wider(names_from = side, values_from = c(rss, maxrss)) |>
    rename(rss_b = !!paste0("rss_", base),  rss_c = !!paste0("rss_", cand),
           mx_b = !!paste0("maxrss_", base), mx_c = !!paste0("maxrss_", cand)) |>
    mutate(rss_ratio = rss_c / rss_b, maxrss_ratio = mx_c / mx_b)
}
```

```{r load}
ab <- load_run(params$ab_file)
has_aa <- !is.null(params$aa_file) && nzchar(params$aa_file) && file.exists(params$aa_file)
aa <- if (has_aa) load_run(params$aa_file) else NULL
ab_stats <- summarise_pair(ab)
aa_stats <- if (has_aa) summarise_pair(aa) else NULL

if (has_aa) {
  aa_join <- aa_stats |> transmute(benchmark, aa_ratio = ratio, own_noise = abs(log(ratio)))
  sigma_g <- mad(log(aa_stats$ratio))
} else {
  aa_join <- tibble(benchmark = ab_stats$benchmark, aa_ratio = NA_real_, own_noise = 0)
  sigma_g <- log(1 + params$min_effect) / 3
}

res <- ab_stats |>
  left_join(aa_join, by = "benchmark") |>
  mutate(own_noise = coalesce(own_noise, 0),
         drift_flag = abs(drift_b - 1) > params$drift_threshold |
                      abs(drift_c - 1) > params$drift_threshold,
         eff = abs(log(ratio)),
         significant = eff > 3 * sigma_g & eff > 2 * own_noise &
                       (lo > 1 | hi < 1) & eff > log(1 + params$min_effect),
         suggestive = !significant & abs(ratio - 1) > params$min_effect &
                      (lo > 1 | hi < 1),
         verdict = case_when(significant & ratio > 1 ~ "regression",
                             significant & ratio < 1 ~ "improvement",
                             TRUE ~ "inconclusive"),
         label = ifelse(drift_flag, paste0(benchmark, " \u2020"), benchmark))

gm <- geomean(res$ratio)
n_reg <- sum(res$verdict == "regression"); n_imp <- sum(res$verdict == "improvement")
regs <- res |> filter(verdict == "regression") |> arrange(desc(ratio))
imps <- res |> filter(verdict == "improvement") |> arrange(ratio)
sugs <- res |> filter(suggestive) |> arrange(desc(eff))

rss_ab <- rss_pair(ab)
if (has_aa) {
  rss_aa <- rss_pair(aa) |> transmute(benchmark, rss_aa = rss_ratio)
  sigma_rss <- mad(log(rss_pair(aa)$rss_ratio), na.rm = TRUE)
} else {
  rss_aa <- tibble(benchmark = rss_ab$benchmark, rss_aa = NA_real_)
  sigma_rss <- log(1.03) / 3
}
rss_res <- rss_ab |>
  left_join(rss_aa, by = "benchmark") |>
  mutate(own = coalesce(abs(log(rss_aa)), 0), eff = abs(log(rss_ratio)),
         flag = eff > pmax(3 * sigma_rss, 2 * own) & abs(rss_ratio - 1) > 0.03)
rss_flagged <- rss_res |> filter(flag) |> arrange(desc(eff))

base_build <- short_build(unlist(ab$desc[ab$sides[1]]))
cand_build <- short_build(unlist(ab$desc[ab$sides[2]]))
```

# Introduction

The data files behind this report are the raw JSON output of
[ruby-bench](https://github.com/ruby/ruby-bench), a benchmark suite and
harness for comparing Ruby builds.

`run_benchmarks.rb` runs every benchmark in
a fresh process for each Ruby executable and records per-iteration wall-clock
times, RSS, and JIT statistics into one JSON file per comparison. The harness
discards the first `r max(ab$meta$n_warmup)` iterations as warmup and then
keeps iterating until a minimum time budget is met, so sample counts differ
between sides. Each process is pinned to one core and runs with ASLR disabled,
as the recorded command line shows:

```{r example-cmd, results="asis"}
if (!is.na(ab$example_cmd)) cat("```\n", ab$example_cmd, "\n```\n", sep = "")
```

## Builds under test

```{r builds-table}
build_rows <- bind_rows(
  parse_build(unlist(ab$desc[ab$sides[1]])) |> mutate(Role = "baseline", .before = 1),
  parse_build(unlist(ab$desc[ab$sides[2]])) |> mutate(Role = "candidate", .before = 1))
if (has_aa) {
  aa_desc <- unique(unlist(aa$desc[aa$sides]))
  build_rows <- bind_rows(build_rows,
    parse_build(aa_desc[1]) |> mutate(Role = "A/A calibration (both sides)", .before = 1))
}
build_rows |>
  transmute(Role, Version = version, Branch = branch, Commit = commit,
            Committed = committed, `Platform / flags` = platform) |>
  kable()
```

Full version strings as recorded in the data files:

```{r build-strings, results="asis"}
cat("```\n")
cat(sprintf("baseline:  %s\n", unlist(ab$desc[ab$sides[1]])))
cat(sprintf("candidate: %s\n", unlist(ab$desc[ab$sides[2]])))
if (has_aa) cat(sprintf("A/A:       %s\n", unique(unlist(aa$desc[aa$sides]))[1]))
cat("```\n")
```

# At a glance

```{r hero, results="asis"}
cat(sprintf(
  '<div class="banner info"><b>%d benchmarks</b> \u00b7 candidate is <b>%s %s</b> overall (geometric mean) \u00b7
   <span class="chip chip-grey">baseline: %s</span>
   <span class="chip chip-blue">candidate: %s</span></div>\n',
  nrow(res), fmt_chg(gm), ifelse(gm > 1, "slower", "faster"), base_build, cand_build))

if (n_reg > 0) {
  w <- regs[1, ]
  cat(sprintf('<div class="banner bad"><b>%d regressions</b> beyond the measured noise floor \u00b7 worst: <b>%s %s</b> (%s \u2192 %s)</div>\n',
              n_reg, w$benchmark, fmt_chg(w$ratio), fmt_ms(w$mu_b), fmt_ms(w$mu_c)))
} else {
  cat('<div class="banner good"><b>Zero regressions.</b> No benchmark is significantly slower on the candidate build.</div>\n')
}
if (n_imp > 0) {
  b <- imps[1, ]
  cat(sprintf('<div class="banner good"><b>%d improvements</b> beyond the noise floor \u00b7 best: <b>%s %s</b> (%s \u2192 %s)</div>\n',
              n_imp, b$benchmark, fmt_chg(b$ratio), fmt_ms(b$mu_b), fmt_ms(b$mu_c)))
}
if (has_aa) {
  cat(sprintf('<div class="banner info">Verdicts are calibrated against an <b>A/A run</b> (identical builds): robust \u03c3 = %.2f%%, significance gate = %s. Within-run 95%% CIs must also exclude zero change.</div>\n',
              100 * sigma_g, fmt_chg(exp(3 * sigma_g))))
} else {
  cat('<div class="banner warn"><b>No A/A calibration file.</b> A fixed 1% threshold replaces the measured noise floor.</div>\n')
}
```

```{r glance-tables, results="asis"}
glance_table <- function(d, cls) {
  max_eff <- max(abs(log(res$ratio[res$significant | res$suggestive])))
  rows <- vapply(seq_len(nrow(d)), function(i) {
    r <- d[i, ]
    w <- max(2, round(100 * abs(log(r$ratio)) / max_eff))
    drift <- ifelse(r$drift_flag,
                    ' <span class="gdrift" title="non-stationary run">\u2020</span>', "")
    aa <- ifelse(is.na(r$aa_ratio), "\u2014",
                 sprintf("\u00b1%.2f%%", 100 * abs(r$aa_ratio - 1)))
    sprintf(paste0(
      '<tr><td class="gname">%s%s</td>',
      '<td class="gchg %s">%s</td>',
      '<td class="gbar"><div class="gfill %s" style="width:%d%%"></div></td>',
      '<td class="gtime">%s \u2192 %s</td>',
      '<td class="gci">%s \u2026 %s</td>',
      '<td class="gaa">%s</td></tr>'),
      r$benchmark, drift,
      ifelse(r$ratio > 1, "pos", "neg"), fmt_chg(r$ratio),
      cls, w,
      fmt_ms(r$mu_b), fmt_ms(r$mu_c),
      fmt_chg(r$lo), fmt_chg(r$hi), aa)
  }, character(1))
  paste0('<table class="gtable"><thead><tr>',
         '<th>Benchmark</th><th class="r">Change</th><th></th>',
         '<th>Baseline \u2192 candidate</th><th>95% CI</th><th>A/A dev</th>',
         '</tr></thead><tbody>', paste(rows, collapse = "\n"), '</tbody></table>')
}
if (nrow(regs) > 0)
  cat('<h2>Significant regressions</h2>\n', glance_table(regs, "reg"), '\n')
if (nrow(imps) > 0)
  cat('<h2>Significant improvements</h2>\n', glance_table(imps, "imp"), '\n')
if (nrow(sugs) > 0)
  cat('<h2>Suggestive (fails the noise gates)</h2>\n', glance_table(sugs, "sug"), '\n')
```

# Data

**A/B comparison file:** `` `r params$ab_file` `` — baseline side
`` `r ab$sides[1]` ``, candidate side `` `r ab$sides[2]` ``.
Ratios are candidate / baseline mean iteration time; above 1 is slower, and
the % change columns show `(ratio - 1)`.

```{r meta-aa, results="asis"}
if (has_aa) {
  cat("**A/A calibration file:** `", params$aa_file, "` \u2014 ",
      "the identical build on both sides measures pipeline noise.\n\n", sep = "")
  aa_desc <- unlist(aa$desc[aa$sides])
  if (length(unique(aa_desc)) != 1) {
    cat('<div class="banner bad"><b>WARNING:</b> the two sides of the A/A file are different builds. The noise calibration is invalid.</div>\n')
  }
}
if (length(unique(unlist(ab$desc[ab$sides]))) == 1) {
  cat('<div class="banner bad"><b>WARNING:</b> both sides of the A/B file report the same build.</div>\n')
}
```

# Methods

- The unit of comparison is the mean post-warmup iteration time per benchmark.
- Each ratio gets a 95% CI by the delta method on the two sample means. The CI
  reflects within-run variance only; between-process variance comes from the
  A/A calibration.
- **Noise floor (A/A):** the spread of A/A log-ratios across benchmarks
  measures the total noise of one invocation pair. The global robust sigma is
  `mad(log ratio)`; each benchmark also gets its own observed A/A deviation.
- **Verdict rule.** A benchmark is a significant regression or improvement
  only when all of these hold:
  1. `|log ratio|` > 3 × global robust sigma from the A/A run;
  2. `|log ratio|` > 2 × that benchmark's own A/A deviation;
  3. the within-run 95% CI excludes 1.0;
  4. the effect exceeds the practical floor of `r percent(params$min_effect)`.
- **Suggestive** effects pass gates 3–4 but fail the noise gates 1–2.
- **Stationarity check:** second-half mean / first-half mean per side. Values
  beyond ±`r percent(params$drift_threshold)` flag the benchmark with †
  (for example, late JIT compilation inside the measurement window).

# Data quality

```{r quality, results="asis"}
drifters <- res |> filter(drift_flag)
range_n <- range(c(res$n_b, res$n_c))
cat(sprintf("Post-warmup sample sizes: %d\u2013%d iterations per side. %d of %d benchmarks drift more than %s between run halves.\n",
            range_n[1], range_n[2], nrow(drifters), nrow(res), percent(params$drift_threshold)))
if (nrow(drifters) > 0) {
  drifters |>
    transmute(Benchmark = benchmark,
              `Baseline drift` = fmt_chg(drift_b),
              `Candidate drift` = fmt_chg(drift_c),
              `n (base / cand)` = sprintf("%d / %d", n_b, n_c)) |>
    kable(caption = "Non-stationary benchmarks \u2014 interpret with caution (\u2020 in charts and tables).") |>
    print()
}
```

# Noise floor from the A/A run

An A/A run feeds the whole pipeline — build, process launch, warmup,
measurement — the *same* Ruby build on both sides. Every deviation from a
ratio of 1.0 in that run is therefore measurement noise: scheduler decisions,
cache and page placement, allocator state, and so on. The table below
summarises how large that noise is across the suite; it is the yardstick that
every A/B verdict in this report is measured against.

```{r noise-table, results="asis"}
if (has_aa) {
  aa_devs <- sort(abs(aa_stats$ratio - 1))
  tibble(Statistic = c("Geometric mean ratio", "Robust sigma of log ratio",
                       "Median |change|", "p90 |change|", "Max |change|"),
         Value = c(fmt_ratio(geomean(aa_stats$ratio)), sprintf("%.2f%%", 100 * sigma_g),
                   percent(median(aa_devs), 0.01), percent(quantile(aa_devs, 0.9), 0.01),
                   paste0(percent(max(aa_devs), 0.01), " (",
                          aa_stats$benchmark[which.max(abs(aa_stats$ratio - 1))], ")"))) |>
    kable() |> print()
} else {
  cat("No A/A file supplied; skipped.\n")
}
```

**What the chart shows.** One point per benchmark: its absolute A/A deviation
|ratio − 1|, ranked from quietest to noisiest, on a logarithmic scale. The
dashed line is the significance gate derived from this distribution.

**The statistics behind it.** The gate is three times a robust standard
deviation of the A/A log-ratios, estimated with the median absolute deviation
(MAD, scaled by 1.4826 so it is comparable to a normal σ). A robust estimator
matters here because the distribution is heavy-tailed: most benchmarks sit
near ±0.1%, but a few allocation- and GC-heavy ones swing by several percent
between identical invocations. A plain standard deviation would let that tail
inflate the gate for everyone. Instead the tail is handled per benchmark: the
verdict rule also requires each A/B effect to exceed **twice the benchmark's
own A/A deviation**, so an invocation-unstable benchmark needs a
correspondingly larger effect before it counts.

```{r noise-plot, fig.height=4.2}
if (has_aa) {
  aa_stats |>
    mutate(dev = abs(ratio - 1)) |>
    arrange(dev) |> mutate(rank = row_number()) |>
    ggplot(aes(rank, pmax(dev, 1e-5))) +
    geom_point(color = C_MARK, size = 1.8) +
    geom_hline(yintercept = exp(3 * sigma_g) - 1, linetype = "dashed",
               color = PAL[["regression"]], linewidth = 0.5) +
    annotate("text", x = 2, y = (exp(3 * sigma_g) - 1) * 1.25, vjust = 0, hjust = 0,
             label = "3 \u00d7 robust sigma (significance gate)",
             color = PAL[["regression"]], size = 3.6) +
    scale_y_log10(labels = percent_format(accuracy = 0.01)) +
    labs(title = "A/A deviations per benchmark (identical builds)",
         subtitle = "Each point is one benchmark. Points above the dashed line are invocation-unstable; their A/B verdicts require 2\u00d7 their own deviation.",
         x = "benchmark, ranked by deviation", y = "|change|, log scale")
}
```

# A/B results

## Significant effects

**What the chart shows.** Each row is a benchmark that passed every
significance gate. The dot is the ratio of mean iteration times
(candidate / baseline) drawn on a logarithmic axis and labelled as a
percentage change — red dots are slower on the candidate, green dots faster.
The horizontal bar through each dot is its 95% confidence interval, the
shaded band around 0% is the ±3σ noise gate from the A/A run, each × is
where the same benchmark landed in the A/A run, and † marks a benchmark whose
run drifted.

**The statistics behind it.** The interval comes from the delta method: the
variance of a ratio of two sample means is approximated from the two relative
variances, and the critical value uses a conservative Student's t with
df = (smaller sample size − 1). This interval captures within-run sampling
error only. Iteration times inside one process are autocorrelated and share
process-level state (memory layout, code placement), so the interval alone
would overstate certainty — that is precisely the gap the A/A gates close.
The logarithmic axis keeps ratios symmetric: ×1.5 (+50%) and ×0.67 (−33%)
sit the same distance from zero, so regressions and improvements can be
compared by eye.

```{r forest-sig, fig.height=max(4, 0.26 * sum(res$significant) + 1.8)}
sig <- res |> filter(significant)
band <- c(exp(-3 * sigma_g), exp(3 * sigma_g))
brks <- c(0.5, 0.67, 0.8, 0.9, 1, 1.1, 1.25, 1.5, 2)
if (nrow(sig) > 0) {
  p <- ggplot(sig, aes(ratio, reorder(label, -ratio), color = verdict)) +
    annotate("rect", xmin = band[1], xmax = band[2], ymin = -Inf, ymax = Inf,
             fill = C_BAND, alpha = 0.55) +
    geom_vline(xintercept = 1, color = C_ZERO) +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, linewidth = 0.7) +
    geom_point(size = 2.4)
  if (has_aa) p <- p + geom_point(aes(x = aa_ratio), shape = 4, color = C_MARK,
                                  size = 1.8, na.rm = TRUE)
  p +
    scale_x_continuous(trans = "log2", breaks = brks, labels = pct_lab) +
    scale_color_manual(values = PAL, guide = "none") +
    labs(title = "Significant changes in iteration time",
         subtitle = "Bars: within-run 95% CI \u00b7 shaded band: 3\u03c3 A/A noise \u00b7 \u00d7 marks: the benchmark's A/A ratio \u00b7 \u2020: non-stationary",
         x = "change vs baseline (log scale) \u2014 right of 0% = slower", y = NULL) +
    theme(axis.text.y = element_text(size = 10.5, color = C_TEXT))
}
```

<details><summary>All `r nrow(res)` benchmarks (forest plot)</summary>

The same encoding as above, now including every benchmark that failed a gate
(grey). Most grey points cluster inside the shaded noise band — the expected
picture for genuine null results.

```{r forest-all, fig.height=max(6, 0.19 * nrow(res) + 1.5)}
p <- ggplot(res, aes(ratio, reorder(label, -ratio), color = verdict)) +
  annotate("rect", xmin = band[1], xmax = band[2], ymin = -Inf, ymax = Inf,
           fill = C_BAND, alpha = 0.55) +
  geom_vline(xintercept = 1, color = C_ZERO) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, linewidth = 0.5) +
  geom_point(size = 1.8)
if (has_aa) p <- p + geom_point(aes(x = aa_ratio), shape = 4, color = C_MARK,
                                size = 1.4, na.rm = TRUE)
p +
  scale_x_continuous(trans = "log2", breaks = brks, labels = pct_lab) +
  scale_color_manual(values = PAL) +
  labs(title = "All benchmarks",
       x = "change vs baseline (log scale) \u2014 right of 0% = slower", y = NULL, color = NULL) +
  theme(axis.text.y = element_text(size = 8))
```

</details>

## Effects versus noise

**What the chart shows.** Each curve is an empirical cumulative distribution
(ECDF): pick a size of change on the x-axis, and the curve height is the
fraction of benchmarks whose deviation is at most that size. The grey curve
is the A/A run; the blue curve is the A/B comparison.

**The statistics behind it.** The A/A curve is the empirical null
distribution — the changes measurement noise produces on its own. If the
candidate build changed nothing, the blue curve would lie on top of the grey
one. Where the blue curve runs to the right of the grey, past the dashed 3σ
gate, the comparison contains effects that noise cannot explain; the
horizontal gap between the curves is how much larger those effects are than
noise at the same quantile. This is the single-picture justification that the
verdicts above are not artifacts of the pipeline.

```{r ecdf, fig.height=4}
ecdf_df <- bind_rows(
  res |> transmute(dev = abs(log(ratio)), run = "A/B (baseline vs candidate)"),
  if (has_aa) aa_stats |> transmute(dev = abs(log(ratio)), run = "A/A (identical builds)"))
ggplot(ecdf_df, aes(pmax(exp(dev) - 1, 1e-5), color = run)) +
  stat_ecdf(linewidth = 0.9) +
  geom_vline(xintercept = exp(3 * sigma_g) - 1, linetype = "dashed", color = C_MARK) +
  scale_x_log10(labels = percent_format(accuracy = 0.01)) +
  scale_color_manual(values = c(C_SIDE_BASE, C_SIDE_CAND)) +
  labs(title = "Where A/B separates from pipeline noise",
       subtitle = "The horizontal gap between the curves right of the dashed line is the set of real effects.",
       x = "|change|, log scale", y = "fraction of benchmarks", color = NULL)
```

## Full results table

```{r verdict-tables, results="asis"}
verdict_table <- function(d) {
  d |>
    arrange(desc(abs(log(ratio)))) |>
    transmute(Benchmark = label,
              Change = sprintf('<span class="%s">%s</span>',
                               ifelse(ratio > 1, "pos", "neg"), fmt_chg(ratio)),
              Ratio = fmt_x(ratio),
              `95% CI` = sprintf("%s \u2026 %s", fmt_chg(lo), fmt_chg(hi)),
              Baseline = fmt_ms(mu_b), Candidate = fmt_ms(mu_c),
              `A/A dev` = ifelse(is.na(aa_ratio), "\u2014",
                                 sprintf("\u00b1%.2f%%", 100 * abs(aa_ratio - 1)))) |>
    kable(escape = FALSE)
}
cat('<details><summary>All benchmarks, sorted by change</summary>\n\n')
print(verdict_table(res))
cat('\n</details>\n')
```

# Iteration traces of the top movers

**What the chart shows.** Every dot is one measured iteration: x is its
position in the run, y its wall-clock time — grey for the baseline process,
blue for the candidate. The panels are the `r params$top_n` largest
significant effects.

**Why it matters statistically.** Everything above compares *means*, and a
mean is only a fair summary of a stationary series. These traces make that
assumption inspectable. Two flat, separated bands are the signature of a
real, stable effect: the vertical separation is the effect size, and the
thickness of each band is the within-run noise the confidence intervals are
built from. Trends, steps, or periodic swings inside a band indicate drift —
late JIT compilation, GC phase changes — which is what the † flag and the
half-run drift check in Data quality detect.

```{r traces, fig.height=7.5}
top <- res |> filter(significant) |> arrange(desc(eff)) |> slice_head(n = params$top_n)
side_pal <- setNames(c(C_SIDE_BASE, C_SIDE_CAND), ab$sides)
side_lab <- setNames(c(paste0(ab$sides[1], " (baseline)"), paste0(ab$sides[2], " (candidate)")),
                     ab$sides)
ab$iters |>
  semi_join(top, by = "benchmark") |>
  mutate(benchmark = factor(benchmark, levels = top$benchmark)) |>
  ggplot(aes(iter, time * 1000, color = side)) +
  geom_point(size = 0.55, alpha = 0.55) +
  facet_wrap(~benchmark, scales = "free", ncol = 2) +
  scale_color_manual(values = side_pal, labels = side_lab,
                     guide = guide_legend(override.aes = list(size = 2.6, alpha = 1))) +
  labs(title = "Per-iteration times, largest significant effects",
       x = "post-warmup iteration", y = "iteration time (ms)", color = NULL)
```

# Memory (post-full-GC RSS)

The harness records `rss` once per process, after a forced full GC — a single
snapshot. Only changes beyond the A/A noise for the same benchmark count.
`maxrss` is the peak from `getrusage`.

```{r rss-summary, results="asis"}
cat(sprintf('<div class="banner %s"><b>%d benchmarks</b> change post-GC RSS beyond the noise gates (3\u03c3 = %s, minimum \u00b13%%).</div>\n',
            ifelse(nrow(rss_flagged) > 0, "warn", "good"), nrow(rss_flagged),
            fmt_chg(exp(3 * sigma_rss))))
if (nrow(rss_flagged) > 0) {
  rss_flagged |>
    transmute(Benchmark = benchmark,
              `RSS change` = sprintf('<span class="%s">%s</span>',
                                     ifelse(rss_ratio > 1, "pos", "neg"), fmt_chg(rss_ratio)),
              `RSS ratio` = fmt_x(rss_ratio),
              `Peak (maxrss)` = ifelse(is.na(maxrss_ratio), "\u2014", fmt_chg(maxrss_ratio)),
              Baseline = sprintf("%.0f MB", rss_b / 2^20),
              Candidate = sprintf("%.0f MB", rss_c / 2^20),
              `A/A dev` = ifelse(is.na(rss_aa), "\u2014",
                                 sprintf("\u00b1%.1f%%", 100 * abs(rss_aa - 1)))) |>
    kable(escape = FALSE) |> print()
}
```

**What the chart shows.** For each flagged benchmark, the filled dot is the
change in resident set size measured once after a forced full GC; the open
circle is the change in peak memory (`getrusage` maxrss) over the whole run.
The axis is logarithmic, matching the time charts.

**The statistics behind it.** Unlike iteration times, there is only one RSS
observation per process, so no confidence interval exists. The verdict leans
entirely on the A/A calibration: a benchmark is flagged only when its RSS
change exceeds both three robust σ of the A/A RSS ratios and twice its own
A/A RSS deviation, with a ±3% practical floor. Post-GC RSS on some benchmarks
is genuinely bimodal across invocations (page-release timing), so single-run
memory verdicts deserve replication before anyone acts on them.

```{r rss-plot, fig.height=max(3, 0.32 * nrow(rss_flagged) + 1.6)}
if (nrow(rss_flagged) > 0) {
  rss_flagged |>
    mutate(direction = ifelse(rss_ratio > 1, "regression", "improvement")) |>
    ggplot(aes(rss_ratio, reorder(benchmark, -rss_ratio), color = direction)) +
    geom_vline(xintercept = 1, color = C_ZERO) +
    geom_point(size = 2.6) +
    geom_point(aes(x = maxrss_ratio), shape = 1, size = 2.6, na.rm = TRUE) +
    scale_x_continuous(trans = "log2", labels = pct_lab) +
    scale_color_manual(values = PAL) +
    labs(title = "RSS changes beyond noise",
         subtitle = "Filled: post-GC RSS \u00b7 open circles: peak (maxrss)",
         x = "memory change vs baseline (log scale) \u2014 right of 0% = more memory",
         y = NULL, color = NULL) +
    theme(axis.text.y = element_text(size = 10.5, color = C_TEXT))
}
```

# Conclusions

```{r conclusions, results="asis"}
top_reg <- regs |> slice_head(n = 3)
top_imp <- imps |> slice_head(n = 3)
cat(sprintf("- The candidate build is **%s %s** than the baseline by geometric mean over %d benchmarks.\n",
            fmt_chg(gm), ifelse(gm > 1, "slower", "faster"), nrow(res)))
if (n_reg > 0) cat(sprintf("- **%d regressions** beyond the noise floor. Largest: %s.\n", n_reg,
  paste(sprintf("%s (%s)", top_reg$benchmark, fmt_chg(top_reg$ratio)), collapse = ", ")))
if (n_imp > 0) cat(sprintf("- **%d improvements** beyond the noise floor. Largest: %s.\n", n_imp,
  paste(sprintf("%s (%s)", top_imp$benchmark, fmt_chg(top_imp$ratio)), collapse = ", ")))
if (nrow(rss_flagged) > 0) cat(sprintf("- **%d memory changes** beyond noise; see the memory section.\n",
                                       nrow(rss_flagged)))
cat("- All verdicts rest on one invocation pair plus the A/A calibration. An effect that must stand alone (a release decision, a regression report) needs replicate invocations of that benchmark.\n")
```

# Reproducing this report

Render with new data by pointing the parameters at other files:

```r
rmarkdown::render(
  "analysis.Rmd",
  params = list(ab_file = "data/new_ab.json", aa_file = "data/new_aa.json"),
  output_file = "analysis-new.html"
)
```

- `ab_file`: ruby-bench JSON; first executable = baseline, second = candidate.
- `aa_file`: the same pipeline with the identical build on both sides.
  Optional; without it a fixed 1% threshold replaces the measured noise floor.
- `min_effect`, `drift_threshold`, `top_n` tune the gates described in Methods.

<details><summary>Session info</summary>

```{r session, echo=FALSE}
sessionInfo()
```

</details>
